---
title: "STAC Test — Panama 2024–2025 — R"
output: html_notebook
---

## Requirements

- **R ≥ 4.2**
- **Packages:** `install.packages(c("rstac", "terra", "httr"))`
- **Network:** HTTP access to `kanopia.org` (STAC API) and `lab.kanopia.org` (COG files)
- **Credentials:** Basic Auth required for COG files — `COG_USER` / `COG_PASS` set in the config cell

# STAC Test — Panama 2024–2025 — R

End-to-end connectivity test for the Kanopia STAC API:

1. Ping the STAC root — confirm the API is reachable
2. Search for all items over **Panama** between **2024-01-01 and 2025-12-31**
3. List all **RGB COG** assets found
4. Render a low-resolution **preview** of the first RGB COG

```{r packages, message=FALSE, warning=FALSE}
# ── Install packages (safe to re-run) ─────────────────────────────────────────
pkgs <- c("rstac", "terra", "httr", "stars")
new  <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new)) install.packages(new)

library(rstac)
library(terra)
library(httr)
library(stars)

`%||%` <- function(a, b) if (is.null(a)) b else a

cat("Packages ready.\n")
```

```{r config}
# ── Configuration ─────────────────────────────────────────────────────────────
STAC_API_URL <- "https://kanopia.org/stac-fastapi-pgstac/api/v1/pgstac/"

# Workshop credentials
COG_USER <- "panama"
COG_PASS <- "panama123"

# Spatial extent: Panama [west, south, east, north]
BBOX <- c(-83.05, 7.20, -77.15, 9.65)

# Temporal extent
DATETIME <- "2024-01-01T00:00:00Z/2025-12-31T23:59:59Z"

# GDAL network settings for remote COG reads
Sys.setenv(
  GDAL_HTTP_TIMEOUT            = "120",
  GDAL_HTTP_MAX_RETRY          = "3",
  GDAL_HTTP_RETRY_DELAY        = "5",
  GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR"
)

add_auth <- function(url, user, pwd) {
  if (grepl("@", sub("^https?://", "", url))) return(url)
  sub("://", paste0("://", user, ":", pwd, "@"), url)
}

cat(sprintf("STAC endpoint : %s\n", STAC_API_URL))
cat(sprintf("Bounding box  : [%s]  (Panama)\n", paste(BBOX, collapse = ", ")))
cat(sprintf("Date range    : %s\n", DATETIME))
```

---
## Step 1 — Ping the STAC API

A simple GET to the root URL confirms the API is reachable and returns its title and version.

```{r ping}
# ── Step 1: Ping STAC root ────────────────────────────────────────────────────
resp    <- httr::GET(STAC_API_URL, httr::timeout(15))
httr::stop_for_status(resp)
catalog <- httr::content(resp, as = "parsed", type = "application/json")

cat(sprintf("  Title   : %s\n", catalog$title        %||% "?"))
cat(sprintf("  Version : %s\n", catalog$stac_version %||% "?"))
cat(sprintf("  ID      : %s\n", catalog$id           %||% "?"))
cat("\nSTAC API is reachable.\n")
```

---
## Step 2 — Search over Panama (2024–2025)

Search the STAC API with a **bounding box** covering Panama and a **date range** of 2024–2025.
No collection filter is applied — all matching items across every project are returned.

```{r stac-search}
# ── Step 2: Search Panama + 2024-2025 ─────────────────────────────────────────
results <- stac(STAC_API_URL, force_version = "1.0.0") |>
  stac_search(
    bbox     = BBOX,
    datetime = DATETIME,
    limit    = 500L
  ) |>
  get_request()

items <- results$features
cat(sprintf("Found %d item(s) over Panama (2024-2025).\n", length(items)))
```

---
## Step 3 — List RGB COG assets

Filter assets to **optimised RGB COGs** (`.cog.tif`, excluding `raw` and `lowres` variants).

```{r list-assets}
# ── Step 3: List RGB COG assets ───────────────────────────────────────────────
rgb_pat     <- "rgb.*\\.cog\\.tif"
exclude_pat <- "raw|lowres|preview"

rgb_assets <- list()

for (item in items) {
  coll   <- item$collection %||% NA_character_
  dt_raw <- item$properties[["datetime"]] %||%
            item$properties[["start_datetime"]] %||% NA_character_
  dt     <- if (!is.na(dt_raw %||% NA)) substr(dt_raw, 1, 10) else NA_character_

  item_assets <- item$assets
  if (!is.list(item_assets)) next

  for (key in names(item_assets)) {
    href <- item_assets[[key]]$href %||% ""
    if ( grepl(rgb_pat,    href, ignore.case = TRUE) &&
        !grepl(exclude_pat, href, ignore.case = TRUE) &&
         grepl("/share/1",  href, fixed = TRUE)) {
      rgb_assets[[length(rgb_assets) + 1]] <- list(
        collection = coll,
        date       = dt,
        href       = href
      )
    }
  }
}

cat(sprintf("Found %d RGB COG asset(s):\n\n", length(rgb_assets)))
if (length(rgb_assets) > 0) {
  df <- do.call(rbind, lapply(rgb_assets, function(x)
    data.frame(collection = x$collection, date = x$date,
               href = x$href, stringsAsFactors = FALSE)))
  print(df, row.names = FALSE)
}
```

---
## Step 4 — COG preview

Open the first RGB COG via `/vsicurl/` (no local download).
`terra::plotRGB()` reads at display resolution via GDAL overviews — no full file download.

```{r cog-preview, fig.width=10, fig.height=8}
# ── Step 4: COG preview of first RGB COG ──────────────────────────────────────
if (length(rgb_assets) == 0) {
  cat("No RGB COG assets found — adjust BBOX or DATETIME.\n")
} else {
  first    <- rgb_assets[[1]]
  url_auth <- add_auth(first$href, COG_USER, COG_PASS)

  cat(sprintf("Collection : %s\n", first$collection))
  cat(sprintf("Date       : %s\n", first$date))
  cat(sprintf("File       : %s\n", basename(first$href)))

  # Open metadata only — no data read yet
  r <- terra::rast(paste0("/vsicurl/", url_auth))
  cat(sprintf("\nDimensions : %d x %d px\n", terra::ncol(r), terra::nrow(r)))
  cat(sprintf("CRS        : %s\n", terra::crs(r, describe = TRUE)$code))
  cat(sprintf("Bands      : %d\n", terra::nlyr(r)))

  # Display at ~512x512 px using stars
  r <- read_stars(
    paste0("/vsicurl/", url_auth),
    RasterIO = list(
      nBufXSize = 512L,
      nBufYSize = 512L,
      bands = 1:3
    )
  )
  plot(r, rgb = 1:3)
}
```
