---
title: "BCI COG Workshop — R"
output: html_notebook
---

## Requirements

- **R ≥ 4.2**
- **Packages:** `install.packages(c("rstac", "terra", "sf"))`
- **Network:** HTTP access to `kanopia.org` (STAC API) and `lab.kanopia.org` (COG files, GeoPackage)
- **Credentials:** Basic Auth required for COG files — set `COG_USER` / `COG_PASS` in the config chunk

# BCI COG Workshop — R

This notebook demonstrates:
1. Querying the Kanopia **STAC API** to find BCI drone imagery by date
2. Reading a tiny window from a **50 GB+ Cloud Optimized GeoTIFF** in ~2 seconds
3. Computing **RGB vegetation indices** (GLI, VARI, GCC) to detect dead trees
4. Comparing the **same tree across two dates** to see change over time

---
**Run this notebook top-to-bottom.** Each chunk builds on the previous one.

```{r packages, message=FALSE, warning=FALSE}
pkgs <- c("rstac", "terra", "sf")
new  <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new)) install.packages(new)

library(rstac)
library(terra)
library(sf)

cat("Packages ready.\n")
```

```{r config}
# ── Configuration ────────────────────────────────────────────────────────────
STAC_API_URL  <- "https://kanopia.org/stac-fastapi-pgstac/api/v1/pgstac/"
COLLECTION_ID <- "2024_bci"

# Workshop credentials
COG_USER <- "panama"
COG_PASS <- "panama123"

# Dead tree to track (from field datasheet — tag surveyed as dead/dying in 2024)
TARGET_TREETAG    <- "6949"
TARGET_POLYGON_ID <- 1041

# Path to field datasheet GeoPackage
GPKG_PATH <- Sys.getenv(
  "GPKG_PATH",
  "https://lab.kanopia.org/share/public/2024-05-14_2024-06-11_fieldDatasheet.gpkg"
)

# GDAL network settings for remote reads
Sys.setenv(
  GDAL_HTTP_TIMEOUT            = "120",
  GDAL_HTTP_MAX_RETRY          = "3",
  GDAL_HTTP_RETRY_DELAY        = "5",
  GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR"
)

cat("STAC endpoint  :", STAC_API_URL,   "\n")
cat("Collection     :", COLLECTION_ID,  "\n")
cat("Target tree tag:", TARGET_TREETAG, "\n")
cat("GeoPackage path:", GPKG_PATH,      "\n")
```

```{r helpers}
# ── Helper functions ──────────────────────────────────────────────────────────

add_auth <- function(url, user, pwd) {
  # Embed Basic Auth into URL: https://user:pwd@host/path
  sub("^(https?://)", paste0("\\1", user, ":", pwd, "@"), url)
}

read_rgb_chip <- function(href, x, y, user, pwd, src_crs = "EPSG:4326", chip_px = 512L) {
  # Read a chip_px x chip_px RGB window from a COG centred on (x, y) in src_crs.
  # Only the small window is downloaded — the full file stays on the server.
  url <- add_auth(href, user, pwd)
  r   <- terra::rast(paste0("/vsicurl/", url))   # reads metadata only (fast)

  # Reproject point to COG CRS using terra
  pt <- terra::vect(matrix(c(x, y), ncol = 2), crs = src_crs)
  pt <- terra::project(pt, terra::crs(r))
  xp <- terra::crds(pt)[1, 1]
  yp <- terra::crds(pt)[1, 2]

  # Window extent: chip_px / 2 pixels on each side
  rx <- terra::res(r)[1]
  ry <- terra::res(r)[2]
  win <- terra::ext(
    xp - chip_px / 2 * rx, xp + chip_px / 2 * rx,
    yp - chip_px / 2 * ry, yp + chip_px / 2 * ry
  )

  terra::crop(r, win)   # only this window is fetched from the server
}

fmt_date <- function(d) {
  d <- as.character(d)
  paste0(substr(d, 1, 4), "-", substr(d, 5, 6), "-", substr(d, 7, 8))
}

rdylgn <- colorRampPalette(c("#d73027", "#fee08b", "#1a9850"))(256)

cat("Helper functions defined.\n")
```

---
## Step 1 — Query the STAC API

We ask the Kanopia STAC API: *"give me all RGB drone mosaics from the `2024_bci` collection"*.

The response is a list of items — each item is one flight date with its COG file URL.

```{r stac-query}
# ── Step 1: Query STAC API ────────────────────────────────────────────────────
pattern <- "\\d{8}_bciwhole_.*rgb\\.cog\\.tif"

results <- stac(STAC_API_URL) |>
  stac_search(collections = COLLECTION_ID, limit = 500) |>
  get_request()

cog_list <- list()
for (item in results$features) {
  for (key in names(item$assets)) {
    href <- item$assets[[key]]$href
    if (!is.null(href) && grepl(pattern, href) && grepl("/share/1", href)) {
      date_str <- regmatches(href, regexpr("\\d{8}", href))
      cog_list[[length(cog_list) + 1]] <- list(
        date    = as.integer(date_str),
        href    = href,
        item_id = item$id
      )
    }
  }
}

cog_list <- cog_list[order(sapply(cog_list, `[[`, "date"))]

cat(sprintf("Found %d BCI whole-island RGB COG assets:\n\n", length(cog_list)))
for (c in cog_list) {
  cat(sprintf("  %s  ->  %s\n", fmt_date(c$date), basename(c$href)))
}
```

---
## Step 2 — Pick a dead tree from the field datasheet

The field GeoPackage contains GPS points from the 2024 BCI tree survey.
We target **tree tag 6949**, which was recorded as dead/dying.

```{r geopkg}
# ── Step 2: Load GeoPackage, extract tree coordinates ────────────────────────
gpkg <- sf::st_read(GPKG_PATH, quiet = TRUE)

matches <- gpkg[
  as.character(gpkg$TreeTag) == TARGET_TREETAG &
  !is.na(gpkg$Polygon_ID) &
  gpkg$Polygon_ID == TARGET_POLYGON_ID, ]

if (nrow(matches) == 0) {
  stop(sprintf("Tree %s / Polygon %s not found in GeoPackage.",
               TARGET_TREETAG, TARGET_POLYGON_ID))
}

geom   <- matches$geometry[[1]]
point  <- if (sf::st_geometry_type(geom) == "POINT") geom else sf::st_centroid(geom)
coords <- sf::st_coordinates(point)

TREE_X   <- coords[1]
TREE_Y   <- coords[2]
TREE_CRS <- sf::st_crs(gpkg)$input   # e.g. "EPSG:32617"

cat("Target tree tag :", TARGET_TREETAG,    "\n")
cat("Polygon ID      :", TARGET_POLYGON_ID, "\n")
cat("GeoPackage CRS  :", TREE_CRS,          "\n")
cat(sprintf("Coordinates     : x=%.2f, y=%.2f\n", TREE_X, TREE_Y))
cat(sprintf("\nWe will read a 512x512 px chip from each of the %d COG dates.\n",
            length(cog_list)))
```

---
## Step 3 — Read one chip (the COG magic moment)

The BCI whole-island RGB COG is **50+ GB**.
We read only the **512x512 pixel window** around our target tree — about **800 KB**.

No download. No waiting. The file stays on the server.

```{r read-chip}
# ── Step 3: Read one RGB chip ─────────────────────────────────────────────────
target_cog <- cog_list[[1]]    # first date — change index to see other dates
label      <- fmt_date(target_cog$date)

cat(sprintf("Reading chip from: %s\n", label))
cat(sprintf("File: %s\n", basename(target_cog$href)))
cat("Fetching from server (only ~800 KB, not the full 50 GB file) ...\n")

# src_crs matches GeoPackage CRS — reprojected directly to COG CRS inside the function
chip <- read_rgb_chip(
  target_cog$href, TREE_X, TREE_Y, COG_USER, COG_PASS,
  src_crs = TREE_CRS
)

terra::plotRGB(chip, r = 1, g = 2, b = 3, stretch = "lin",
               main = paste0("RGB chip - Tree ", TARGET_TREETAG, " - ", label))

cat(sprintf("\nDone. Chip size: %dx%d px - read in seconds from a 50 GB file.\n",
            ncol(chip), nrow(chip)))
```

---
## Step 4 — Compute RGB vegetation indices

These three indices use only **Red, Green, Blue** bands — perfect for drone RGB data.

| Index | Formula | What it measures |
|-------|---------|------------------|
| **GLI** — Green Leaf Index | `(2G - R - B) / (2G + R + B)` | Greenness; high = healthy |
| **VARI** — Visible Atmospherically Resistant Index | `(G - R) / (G + R - B)` | Robust greenness, less sensitive to haze |
| **GCC** — Green Chromatic Coordinate | `G / (R + G + B)` | Canopy phenology; tracks seasonal greenup/senescence |

**Dead trees → low GLI, low VARI, low GCC** (red in the RdYlGn palette).

```{r compute-indices}
# ── Step 4: Compute GLI, VARI, GCC directly on terra SpatRasters ─────────────
eps <- 1e-6
R   <- chip[[1]]; G <- chip[[2]]; B <- chip[[3]]

GLI  <- (2*G - R - B) / (2*G + R + B + eps)
VARI <- clamp((G - R) / (G + R - B + eps), -1.5, 1.5)
GCC  <- G / (R + G + B + eps)

cat(sprintf("GLI  range: [%.3f, %.3f]  (healthy forest ~+0.1 to +0.4)\n",
            global(GLI,  "min", na.rm = TRUE)[[1]],
            global(GLI,  "max", na.rm = TRUE)[[1]]))
cat(sprintf("VARI range: [%.3f, %.3f]  (healthy ~0.0 to +0.5)\n",
            global(VARI, "min", na.rm = TRUE)[[1]],
            global(VARI, "max", na.rm = TRUE)[[1]]))
cat(sprintf("GCC  range: [%.3f, %.3f]  (healthy ~0.35 to 0.45)\n",
            global(GCC,  "min", na.rm = TRUE)[[1]],
            global(GCC,  "max", na.rm = TRUE)[[1]]))
```

```{r plot-indices, fig.width=20, fig.height=5}
# ── Step 5: Side-by-side visualization ───────────────────────────────────────
par(mfrow = c(1, 4), mar = c(1, 1, 3, 2))

terra::plotRGB(chip, r = 1, g = 2, b = 3, stretch = "lin", axes = FALSE,
               main = paste0("RGB\n", label))

plot(GLI,  col = rdylgn, range = c(-0.5, 0.5),  axes = FALSE, legend = TRUE,
     main = "GLI\n(Green Leaf Index)")
plot(VARI, col = rdylgn, range = c(-0.5, 0.8),  axes = FALSE, legend = TRUE,
     main = "VARI\n(Atmospherically Resistant)")
plot(GCC,  col = rdylgn, range = c(0.28, 0.45), axes = FALSE, legend = TRUE,
     main = "GCC\n(Green Chromatic Coord.)")

mtext(
  paste0("BCI - Tree ", TARGET_TREETAG, " - ", label,
         "\nRed = low greenness (dead/stressed)   Green = healthy canopy"),
  outer = TRUE, line = -1.5, cex = 0.9, font = 2
)
```

---
## Step 6 — Track the dead tree across two dates

We now read the **same chip** from two dates and compare GLI maps.

**Healthy canopy** stays green (high GLI).
**Dead or dying tree** turns red (low GLI) between the two dates.

```{r twodate, fig.width=12, fig.height=12}
# ── Step 6: Two-date GLI comparison ──────────────────────────────────────────
date_early <- cog_list[[1]]
date_late  <- cog_list[[length(cog_list)]]

cat(sprintf("Early date : %s\n", fmt_date(date_early$date)))
cat(sprintf("Late  date : %s\n", fmt_date(date_late$date)))
cat("Reading both chips...\n")

chip_e <- read_rgb_chip(date_early$href, TREE_X, TREE_Y, COG_USER, COG_PASS,
                        src_crs = TREE_CRS)
chip_l <- read_rgb_chip(date_late$href,  TREE_X, TREE_Y, COG_USER, COG_PASS,
                        src_crs = TREE_CRS)

gli_e <- (2*chip_e[[2]] - chip_e[[1]] - chip_e[[3]]) /
         (2*chip_e[[2]] + chip_e[[1]] + chip_e[[3]] + 1e-6)
gli_l <- (2*chip_l[[2]] - chip_l[[1]] - chip_l[[3]]) /
         (2*chip_l[[2]] + chip_l[[1]] + chip_l[[3]] + 1e-6)

par(mfrow = c(2, 2), mar = c(1, 1, 3, 2))

terra::plotRGB(chip_e, r = 1, g = 2, b = 3, stretch = "lin", axes = FALSE,
               main = paste0("RGB - ", fmt_date(date_early$date)))
terra::plotRGB(chip_l, r = 1, g = 2, b = 3, stretch = "lin", axes = FALSE,
               main = paste0("RGB - ", fmt_date(date_late$date)))

plot(gli_e, col = rdylgn, range = c(-0.5, 0.5), axes = FALSE, legend = TRUE,
     main = paste0("GLI - ", fmt_date(date_early$date)))
plot(gli_l, col = rdylgn, range = c(-0.5, 0.5), axes = FALSE, legend = TRUE,
     main = paste0("GLI - ", fmt_date(date_late$date)))

mtext(
  paste0("Tree ", TARGET_TREETAG, " - Temporal Change in GLI\n",
         "Green = healthy   Red = dead/stressed"),
  outer = TRUE, line = -1.5, cex = 0.9, font = 2
)
```

---
## What just happened?

| Step | What we did | Data transferred |
|------|-------------|------------------|
| STAC query | Found all BCI RGB COGs for 2024 | ~15 KB |
| COG chip read (x2) | Read 512x512 px window centred on tree | ~800 KB each |
| Index computation | GLI, VARI, GCC | 0 bytes (local math) |

**The BCI whole-island RGB COG is 50+ GB.**
We transferred less than 2 MB total — about **0.004% of the file**.

This is the power of **Cloud Optimized GeoTIFF + STAC**:
- **STAC** tells you *which* file and *where* to find it
- **COG** lets you read *only the part you need*, from anywhere, with any tool
