---
title: "STAC Search by Spatial & Temporal Extent"
output: html_notebook
---

## Requirements

- **R ≥ 4.2**
- **Packages:** `install.packages(c("rstac", "dplyr", "tidyr", "ggplot2"))`
- **Network:** HTTP access to `kanopia.org` (STAC API)

# STAC Search by Spatial & Temporal Extent

Query the Kanopia STAC API using only a **bounding box** and a **date range** — no collection filter.

Use case: *find all RGB, DSM, and LAZ assets over global tropical forests between 2019 and 2025.*

---
**Steps**
1. Search the STAC API — bbox + datetime, no collection filter
2. Filter to RGB / DSM / COPC LAZ (exclude raw) and summarise by project

```{r packages, message=FALSE, warning=FALSE}
pkgs <- c("rstac", "dplyr", "tidyr", "ggplot2")
new  <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new)) install.packages(new)

library(rstac)
library(dplyr)
library(tidyr)
library(ggplot2)

# Null-coalescing operator
`%||%` <- function(a, b) if (is.null(a)) b else a

cat("Packages ready.\n")
```

```{r config}
# ── Configuration ─────────────────────────────────────────────────────────────
STAC_API_URL <- "https://kanopia.org/stac-fastapi-pgstac/api/v1/pgstac/"

# Workshop credentials
COG_USER <- "panama"
COG_PASS <- "panama123"

# Spatial extent: global tropical belt (Tropic of Cancer to Tropic of Capricorn)
# Covers all tropical forests worldwide: Amazon, Congo, SE Asia, Central America, etc.
BBOX <- c(-180, -23.5, 180, 23.5)

# Temporal extent: ISO-8601 interval
DATETIME <- "2019-01-01T00:00:00Z/2025-12-31T23:59:59Z"

MAX_ITEMS <- 9999L

cat(sprintf("STAC endpoint : %s\n", STAC_API_URL))
cat(sprintf("Bounding box  : [%s]  (global tropical belt)\n",
            paste(BBOX, collapse = ", ")))
cat(sprintf("Date range    : %s\n", DATETIME))
```

---
## Step 1 — Search by spatial + temporal extent

We pass `bbox` and `datetime` to the STAC search — **no `collections` filter**.
The API returns every item whose spatial footprint intersects the bounding box
and whose acquisition date falls within the date range.

```{r stac-search}
# ── Step 1: STAC search — bbox + datetime, no collection filter ───────────────
results <- stac(STAC_API_URL) |>
  stac_search(
    bbox     = BBOX,
    datetime = DATETIME,
    limit    = MAX_ITEMS
  ) |>
  get_request()

items <- results$features
cat(sprintf("Found %d item(s) matching the spatial + temporal extent.\n", length(items)))
```

---
## Step 2 — Summary by project (collection)

Filter to **RGB**, **DSM**, and **LAZ** assets (exclude raw).
Count assets per collection, with a grand total row.

```{r filter-assets}
# ── Step 2: Filter assets and summarise by project ────────────────────────────
include_pat <- "rgb.*\\.cog\\.tif|dsm.*\\.cog\\.tif|copc.*\\.laz"
exclude_pat <- "raw|preview"

assets_rows <- list()

for (item in items) {
  coll   <- item$collection %||% item$properties[["collection"]] %||% NA_character_
  dt_raw <- item$properties[["datetime"]] %||%
            item$properties[["start_datetime"]] %||% NA_character_
  dt     <- if (!is.na(dt_raw %||% NA)) substr(dt_raw, 1, 10) else NA_character_

  item_assets <- item$assets
  if (!is.list(item_assets)) next

  for (key in names(item_assets)) {
    href  <- item_assets[[key]]$href %||% ""
    label <- paste(key, href)

    if (!grepl(include_pat, label, ignore.case = TRUE)) next
    if ( grepl(exclude_pat, label, ignore.case = TRUE)) next

    asset_type <- dplyr::case_when(
      grepl("copc|\\.laz$", label, ignore.case = TRUE) ~ "LAZ",
      grepl("dsm",          label, ignore.case = TRUE) ~ "DSM",
      TRUE                                              ~ "RGB"
    )

    assets_rows[[length(assets_rows) + 1]] <- data.frame(
      collection = coll,
      datetime   = dt,
      asset_type = asset_type,
      stringsAsFactors = FALSE
    )
  }
}

if (length(assets_rows) == 0) {
  cat("No assets found — adjust BBOX or DATETIME and re-run.\n")
} else {
  assets_df <- dplyr::bind_rows(assets_rows)

  summary_tbl <- assets_df |>
    dplyr::count(collection, asset_type) |>
    tidyr::pivot_wider(
      names_from  = asset_type,
      values_from = n,
      values_fill = 0L
    ) |>
    dplyr::mutate(
      RGB   = dplyr::coalesce(RGB,   0L),
      DSM   = dplyr::coalesce(DSM,   0L),
      LAZ   = dplyr::coalesce(LAZ,   0L),
      Total = RGB + DSM + LAZ
    ) |>
    dplyr::arrange(dplyr::desc(Total))

  date_range <- assets_df |>
    dplyr::group_by(collection) |>
    dplyr::summarise(
      first_date = min(datetime, na.rm = TRUE),
      last_date  = max(datetime, na.rm = TRUE),
      .groups    = "drop"
    )

  summary_tbl <- dplyr::left_join(summary_tbl, date_range, by = "collection")

  # Grand total row
  total_row <- summary_tbl |>
    dplyr::summarise(
      collection = "TOTAL",
      RGB   = sum(RGB),
      DSM   = sum(DSM),
      LAZ   = sum(LAZ),
      Total = sum(Total),
      first_date = min(first_date, na.rm = TRUE),
      last_date  = max(last_date,  na.rm = TRUE)
    )

  cat(strrep("=", 60), "\n")
  cat(sprintf("  Assets: %d  |  Collections: %d\n",
              nrow(assets_df), dplyr::n_distinct(assets_df$collection)))
  cat(strrep("=", 60), "\n")
  print(dplyr::bind_rows(summary_tbl, total_row), n = Inf)
}
```

```{r bar-chart, fig.width=12, fig.height=5}
# ── Bar chart ─────────────────────────────────────────────────────────────────
if (exists("summary_tbl") && nrow(summary_tbl) > 0) {
  plot_df <- summary_tbl |>
    dplyr::select(collection, RGB, DSM, LAZ) |>
    tidyr::pivot_longer(
      cols      = c(RGB, DSM, LAZ),
      names_to  = "type",
      values_to = "n"
    ) |>
    dplyr::mutate(type = factor(type, levels = c("RGB", "DSM", "LAZ")))

  ggplot(plot_df, aes(x = collection, y = n, fill = type)) +
    geom_col(position = "dodge", width = 0.75) +
    scale_fill_manual(
      values = c(RGB = "#4caf50", DSM = "#2196f3", LAZ = "#9c27b0")
    ) +
    labs(
      title = paste0(
        "Geospatial assets by project - global tropical belt (+-23.5 deg)\n",
        substr(DATETIME, 1, 10), " -> ", substr(DATETIME, 22, 31)
      ),
      x    = "Collection (project)",
      y    = "Number of assets",
      fill = "Asset type"
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```
