---
title: "List COGs for collection `2024_bci` — R"
output: html_notebook
---

## Requirements

- **R ≥ 4.2**
- **Packages:** `install.packages(c("rstac", "terra", "sf"))`
- **Network:** HTTP access to `kanopia.org` (STAC API) and `lab.kanopia.org` (COG files, GeoPackage)
- **Credentials:** Basic Auth required for COG files — set `COG_USER` / `COG_PASS` in the config chunk

# List COGs for collection `2024_bci` — R

This notebook queries the Kanopia STAC API, lists all COG assets matching the `2024_bci` collection and the `*bciwhole*rgb.cog.tif` pattern, then extracts a 512×512 pixel PNG chip centred on a target tree for each flight date.

```{r packages, message=FALSE, warning=FALSE}
pkgs <- c("rstac", "terra", "sf")
new  <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new)) install.packages(new)

library(rstac)
library(terra)
library(sf)

cat("Packages ready.\n")
```

```{r config}
# ── Configuration ─────────────────────────────────────────────────────────────
STAC_API_URL  <- "https://kanopia.org/stac-fastapi-pgstac/api/v1/pgstac/"
COLLECTION_ID <- "2024_bci"

# Workshop credentials
COG_USER <- "panama"
COG_PASS <- "panama123"

# Field datasheet GeoPackage
GPKG_PATH         <- "https://lab.kanopia.org/share/public/2024-05-14_2024-06-11_fieldDatasheet.gpkg"
TARGET_TREETAG    <- "6949"
TARGET_POLYGON_ID <- 1041

OUTPUT_DIR <- "bci_cog_snips"

# GDAL network settings for remote reads
Sys.setenv(
  GDAL_HTTP_TIMEOUT            = "120",
  GDAL_HTTP_MAX_RETRY          = "3",
  GDAL_HTTP_RETRY_DELAY        = "5",
  GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR"
)

cat("STAC endpoint:", STAC_API_URL, "\n")
cat("Collection   :", COLLECTION_ID, "\n")
```

```{r list-cogs}
# ── List all bciwhole RGB COGs ────────────────────────────────────────────────
pattern <- "\\d{8}_bciwhole_.*rgb\\.cog\\.tif"

results <- stac(STAC_API_URL) |>
  stac_search(collections = COLLECTION_ID, limit = 500) |>
  get_request()

cog_list <- list()
for (item in results$features) {
  for (key in names(item$assets)) {
    href <- item$assets[[key]]$href
    if (!is.null(href) && grepl(pattern, href)) {
      date_str <- regmatches(href, regexpr("\\d{8}", href))
      cog_list[[length(cog_list) + 1]] <- list(
        item_id   = item$id,
        asset_key = key,
        href      = href,
        date      = as.integer(date_str)
      )
    }
  }
}

cog_list <- cog_list[order(sapply(cog_list, `[[`, "date"))]

cat(sprintf("Found %d matching COG assets for collection %s\n\n",
            length(cog_list), COLLECTION_ID))
for (i in seq_along(cog_list)) {
  c <- cog_list[[i]]
  cat(sprintf("%03d: %d | %s | %s | %s\n",
              i, c$date, c$item_id, c$asset_key, c$href))
}
```

```{r extract-chips}
# ── Extract 512x512 PNG chips centred on target tree ──────────────────────────

add_auth <- function(url, user, pwd) {
  # Embed Basic Auth into URL: https://user:pwd@host/path
  sub("^(https?://)", paste0("\\1", user, ":", pwd, "@"), url)
}

# Clean output directory before writing new chips
if (dir.exists(OUTPUT_DIR)) {
  unlink(OUTPUT_DIR, recursive = TRUE)
  cat(sprintf("Cleaned existing directory: %s\n", OUTPUT_DIR))
}
dir.create(OUTPUT_DIR, recursive = TRUE)

# Load GeoPackage and find target tree point
cat("Loading geopackage:", GPKG_PATH, "\n")
gpkg <- sf::st_read(paste0("/vsicurl/", GPKG_PATH), quiet = TRUE)
cat("Geopackage layer rows:", nrow(gpkg), "\n")

matches <- gpkg[which(
  as.character(gpkg$TreeTag) == TARGET_TREETAG &
  !is.na(gpkg$Polygon_ID) &
  as.numeric(gpkg$Polygon_ID) == as.numeric(TARGET_POLYGON_ID)
), ]

if (nrow(matches) == 0) {
  stop("No matching row found for treetag/polygon_id in geopackage.")
}

geom   <- matches$geometry[[1]]
point  <- if (sf::st_geometry_type(geom) == "POINT") geom else sf::st_centroid(geom)
coords <- sf::st_coordinates(point)

tree_x   <- coords[1]
tree_y   <- coords[2]
tree_crs <- sf::st_crs(gpkg)$input   # e.g. "EPSG:32617"

n <- length(cog_list)
cat(sprintf("Found point at x=%.6f, y=%.6f (CRS: %s). Processing %d COGs...\n",
            tree_x, tree_y, tree_crs, n))

for (idx in seq_along(cog_list)) {
  c    <- cog_list[[idx]]
  href <- c$href

  if (grepl("request-access", href)) {
    cat(sprintf("[%d/%d] skipped (request-access): %s\n", idx, n, href))
    next
  }

  href_auth <- add_auth(href, COG_USER, COG_PASS)

  # Name PNG after the COG filename (without .tif) + suffix — matches Python version
  cog_stem <- tools::file_path_sans_ext(basename(href))   # e.g. '20240611_bciwhole_rx1rii_rgb.cog'
  out_png  <- file.path(OUTPUT_DIR, paste0(cog_stem, "_512x512_chip.png"))

  # Open raster — reads metadata only at this point (fast)
  r <- tryCatch(
    terra::rast(paste0("/vsicurl/", href_auth)),
    error = function(e) {
      cat(sprintf("[%d/%d] error opening: %s\n", idx, n, href)); NULL
    }
  )
  if (is.null(r)) next

  # Reproject tree point to COG CRS using terra
  pt <- terra::vect(matrix(c(tree_x, tree_y), ncol = 2), crs = tree_crs)
  pt <- terra::project(pt, terra::crs(r))
  xp <- terra::crds(pt)[1, 1]
  yp <- terra::crds(pt)[1, 2]

  # Build 512x512 px window extent — src.index() returns (row, col), careful with order
  rx <- terra::res(r)[1]
  ry <- terra::res(r)[2]

  rext    <- terra::ext(r)
  xmin_w  <- max(xp - 256 * rx, terra::xmin(rext))
  xmax_w  <- min(xp + 256 * rx, terra::xmax(rext))
  ymin_w  <- max(yp - 256 * ry, terra::ymin(rext))
  ymax_w  <- min(yp + 256 * ry, terra::ymax(rext))

  if (xmin_w >= xmax_w || ymin_w >= ymax_w) {
    cat(sprintf("[%d/%d] skipped (window out of bounds): %s\n", idx, n, cog_stem))
    next
  }

  win  <- terra::ext(xmin_w, xmax_w, ymin_w, ymax_w)

  # Read only this small window — COG magic: ~800 KB from a 50 GB file
  nlyr_use <- min(terra::nlyr(r), 3L)
  chip     <- terra::crop(r[[seq_len(nlyr_use)]], win)

  # Write PNG via grDevices — no extra packages required
  png(out_png, width = ncol(chip), height = nrow(chip))
  par(mar = c(0, 0, 0, 0))
  terra::plotRGB(chip, r = 1, g = 2, b = 3, stretch = "lin", axes = FALSE)
  dev.off()

  cat(sprintf("[%d/%d] wrote %s\n", idx, n, out_png))
}

cat(sprintf("Done: extracted PNGs in %s\n", OUTPUT_DIR))
```

## Notes
- If no entries are found, ensure the `COLLECTION_ID` exists and the API endpoint is available.
- You can customize `limit`, the regex pattern, or the chip size (currently 512x512).
